---
title: "Tracking input data with a manifest"
author: "Alex M. Chubaty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tracking input data with a manifest}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`workflowtools` ships an input-data manifest API for projects that consume
multiple datasets from heterogeneous sources (BC Data Catalogue, NRCan, ERA5,
hand-curated files). The manifest is a single JSON file that records, for each
dataset:

- its **source** (catalogue URL, query, retrieval timestamp),
- its **vintage** (publisher's version or last-modified date),
- a **local pointer** plus content **hash**,
- and a structured **citation** (BibTeX key, full entry, prose).

Downstream tooling (e.g. a per-PDF provenance appendix or a `data-sources.bib`
sidecar) reads the manifest to populate itself, so the project's input
documentation stays in sync with what was actually fetched.

This vignette walks through the typical lifecycle.

```{r setup}
library(workflowtools)
```

## 1. Where does the manifest live?

Convention: a single JSON file at `data/input_manifest.json`, alongside the
project's `data/` directory. The default arguments to every manifest function
point there, so most calls don't need to mention the path.

For this vignette we'll use a temp file.

```{r}
manifest_path <- file.path(tempdir(), "input_manifest.json")
```

## 2. Register the first input -- a hand-written record

For sources without a dedicated extractor (a USB-stick drop, an internal
reference layer, an ERA5 batch download), build a record manually with
`input_manifest_record()`:

```{r}
nfdb <- input_manifest_record(
  id = "nfdb",
  name = "Canadian National Fire Database (NFDB) -- Agency Fire Data",
  source = list(
    type = "http_download",
    url = "https://cwfis.cfs.nrcan.gc.ca/en_CA/nfdb"
  ),
  local_path = "data/processed/nfdb_poly.gpkg",
  version_or_vintage = "2024-12-01",
  license = "OGL-Canada-2.0",
  citation = list(
    bibtex_key = "CFS:NFDB",
    formatted_text = "Canadian Forest Service (2024). Canadian National Fire Database -- Agency Fire Data."
  )
)
register_input(nfdb, manifest_path)
```

`register_input()` is idempotent and keyed by `id`: re-running with an updated
`nfdb` record replaces the existing entry rather than duplicating it.

```{r}
nfdb$version_or_vintage <- "2025-01-15"
register_input(nfdb, manifest_path)

read_input_manifest(manifest_path)[[1]]$version_or_vintage
```

## 3. Register a BC Data Catalogue record automatically

For datasets that flow through the BC Data Catalogue, `metadata_bcdata()`
fetches the catalogue record, extracts the relevant fields, caches them, and
returns a manifest-ready record:

```{r, eval = FALSE}
vri <- metadata_bcdata("2ebb35d8-c82f-4a17-9c96-612ac3532d55")
register_input(vri, manifest_path)
```

The cache (default location:
`tools::R_user_dir("workflowtools", which = "cache")/metadata/`) means a
second call returns immediately without hitting the network. Cache TTL is 30
days by default; override with `metadata_bcdata(..., ttl_days = 90)` or
`options(workflowtools.metadata_cache.ttl_days = 90)`. Pass
`refresh = TRUE` to force a fetch.

Behind the scenes `metadata_bcdata()` calls
`bcdata::bcdc_get_record()` and dispatches its result to the S3 method
`extract_metadata.bcdc_record()`. You can call those directly if you want
to short-circuit the cache:

```{r, eval = FALSE}
bcdata::bcdc_get_record("...") |> extract_metadata()
```

## 4. Register a raw URL with the fallback method

For an arbitrary URL with no dedicated metadata API,
`extract_metadata.character()` does a best-effort HTTP HEAD to populate the
record's vintage from `Last-Modified`:

```{r, eval = FALSE}
era5_doc <- extract_metadata(
  "https://cds.climate.copernicus.eu/cdsapp#!/dataset/reanalysis-era5-single-levels",
  id = "era5-single-levels"
)
register_input(era5_doc, manifest_path)
```

Pass `head_lookup = FALSE` to skip the HEAD request (useful when offline or
when the server is known to reject HEAD).

## 5. Extending: a custom S3 method

If you have a source not covered by the built-in methods, define an S3 method
for `extract_metadata`. Here's a sketch for a hypothetical NRCan FBP fuel-type
record:

```{r, eval = FALSE}
extract_metadata.nrcan_fbp_record <- function(x, id = NULL, ...) {
  input_manifest_record(
    id        = id %||% "cffdrs-fbp-fuels-2024",
    name      = x$dataset_title,
    source    = list(type = "nrcan_fbp", url = x$portal_url),
    local_path = file.path("data", x$file_name),
    license   = x$license,
    citation  = list(
      bibtex_key = sprintf("NRCan:%s", x$version_year),
      formatted_text = sprintf("NRCan (%s). %s.", x$version_year, x$dataset_title)
    )
  )
}
```

Subsequent calls to `extract_metadata(record)` dispatch on
`class(record)[1L]` and your method takes over.

## 6. Sync citations to a sidecar BibTeX file

The bridge to a project's bibliography is `sync_manifest_to_bibtex()`:
a one-way, conservative exporter that writes eligible manifest citations to a
sidecar `.bib`, deduplicated against the project's hand-curated bibliography.

```{r}
references_bib <- file.path(tempdir(), "references.bib")
out_bib        <- file.path(tempdir(), "data-sources.bib")

## A hand-curated references.bib that already contains an entry for one of
## the manifest records; the sync should skip it (and warn) to keep the
## curated bib the source of truth for that key.
writeLines("@misc{CFS:NFDB,\n  title = {Curated NFDB entry},\n}", references_bib)

result <- sync_manifest_to_bibtex(
  manifest       = manifest_path,
  out            = out_bib,
  references_bib = references_bib
)
result
```

The collision warning is a deliberate prompt: either delete the duplicate
entry from `references.bib` (letting the sync own the key from now on) or
mark the manifest record with `citation = list(..., external = TRUE)` to
silence the warning and keep the curated entry authoritative.

Records with `citation = list(external = TRUE)` are skipped silently:

```{r}
nfdb$citation$external <- TRUE
register_input(nfdb, manifest_path)
sync_manifest_to_bibtex(manifest_path, out_bib, references_bib)
```

## 7. Reading the manifest from downstream consumers

Provenance appendices, audit tables, and bibliography pre-renders all consume
the manifest via `read_input_manifest()`:

```{r}
recs <- read_input_manifest(manifest_path)
data.frame(
  id      = vapply(recs, `[[`, character(1L), "id"),
  vintage = vapply(recs, function(r) r$version_or_vintage %||% "", character(1L))
)
```

For prose output, [`citation_text()`] and [`as_bibentry()`] handle records
one at a time.

## 8. Recommended project wiring

A typical project arranges things like this:

1. Each input-fetch target in your pipeline (`{targets}`, drake, Make, ...)
   calls `register_input(extract_metadata(source_obj), manifest_path)` after
   the fetch succeeds.
2. A final target calls `sync_manifest_to_bibtex()` once per build.
3. The qmd / Rmd report's bibliography lists both the curated and
   auto-generated `.bib` files.
4. A "provenance appendix" R chunk reads the manifest and renders a kable of
   inputs with their vintages and citations.

The result: every report ships with a self-describing record of where its
data came from -- without a hand-maintained input list to keep in sync.

## See also

- [`input_manifest_schema()`] for the canonical field list.
- [`extract_metadata()`] for the S3 generic + the methods shipped with
  this package (`character`, `bcdc_record`).
- [`sync_manifest_to_bibtex()`] for the dedup rules.
